Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-08-02 paper/paper.tex
Fix spelling errors

Nervous systems process and transmit information using electrically excitable membranes. Conductance-based models are a powerful biophysical simplification of an electrically excitable compartment in a neuron (\cite{hodgkinQuantitativeDescriptionMembrane1952a}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small circuits (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). These models provide an approachable framework for understanding many salient principles of neuroscience, since they explicitly model cell membranes and ion channels as electrical components in a circuit. However, several challenges remain in working with biophysically-detailed conductance-based neuron models. First, these models can be high-dimensional with many nonlinear differential equations, with several parameters. Second, many or all equations in these models can be strongly coupled through variables like the membrane potential.  In multi-compartment models of spatially extended neurons, membrane potentials in every compartment are coupled to each other. Finally, the choice of programming language used to implement the model imposes tradeoffs in designing and using neuron and network simulators: simulators written in languages like "C++", "FORTRAN" or "hoc" can integrate equations quickly, but often lack the ease-of-use and interoperability of those written in scientific programming languages like "Python", "Julia" or "MATLAB".

@@

To overcome these design limitations, we have developed a novel automatic type system, "cpplab", which binds "MATLAB" code to "C++" header files. This architecture automatically creates objects in "MATLAB" that represent the underlying object-oriented "C++" code, without having to re-specify these objects in "MATLAB". In this paper, we introduce "xolotl", an implementation of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of "MATLAB" code using a hierarchical and intuitive syntax. Since models in the "MATLAB" workspace are automatically linked to models in the "C++" implementation, configuring these objects in "MATLAB" transparently configures the underlying "C++" objects. "xolotl" comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a graphical user interface (GUI) for real-time manipulation of model parameters. The GUI permits real-time manipulation of any network parameter and an immediate visualization of the effect of that parameter on arbitrary dynamical features of the model. The ease-of-use of these tools lends them to pedagogical applications and the rapid prototyping of models. This tool aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

>>>>

Nervous systems process and transmit information using electrically excitable membranes. Conductance-based models are a powerful biophysical simplification of an electrically excitable compartment in a neuron (\cite{hodgkinQuantitativeDescriptionMembrane1952a}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small circuits (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). These models provide an approachable framework for understanding many salient principles of neuroscience, since they explicitly model cell membranes and ion channels as electrical components in a circuit. However, several challenges remain in working with biophysically-detailed conductance-based neuron models. First, these models can be high-dimensional with many nonlinear differential equations, each with several parameters. Second, many or all equations in these models can be strongly coupled through variables like the membrane potential.  In multi-compartment models of spatially extended neurons, membrane potentials in every compartment are coupled to each other. Finally, the choice of programming language used to implement the model imposes tradeoffs in designing and using neuron and network simulators: simulators written in languages like "C++", "FORTRAN" or "hoc" can integrate equations quickly, but often lack the ease-of-use and interoperability of those written in scientific programming languages like "Python", "Julia" or "MATLAB".

@@

To overcome these design limitations, we have developed a novel automatic type system, "cpplab", which binds "MATLAB" code to "C++" header files. This architecture automatically creates objects in "MATLAB" that represent the underlying object-oriented "C++" code, without having to re-specify these objects in "MATLAB". In this paper, we introduce "xolotl", an implementation of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of "MATLAB" code using a hierarchical and intuitive syntax. Since models in the "MATLAB" workspace are automatically linked to models in the "C++" implementation, configuring these objects in "MATLAB" transparently configures the underlying "C++" objects. "xolotl" comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. The GUI permits real-time manipulation of any network parameter and an immediate visualization of the effect of that parameter on arbitrary dynamical features of the model. The ease-of-use of these tools lends them to pedagogical applications and the rapid prototyping of models. This tool aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.
